

Reproducible Research : Course Project 1
==========================

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "images/",
  cache = TRUE
)

```

## Loading and preprocessing the data

```{r Loading and Processing Data}
url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url,destfile = "activity.zip")
unzip("./activity.zip", exdir = "./activity")
act<-read.csv("./activity/activity.csv",na.strings = "NA")
##Transforming the Date field
act$date<-as.Date(as.character(act$date),"%Y-%m-%d")
```

## What is mean total number of steps taken per day?

```{r Avg Steps taken per day}

## Using tapply function to find total no of steps per day
stepsum<-with(act,tapply(steps, date, sum,na.rm=T))
dates<-names(stepsum)
totalSteps<-as.numeric(stepsum[names(stepsum)])
steptable<-data.frame(dates,totalSteps)

hist(steptable$totalSteps,main = "Histogram of total steps taken per day",xlab = "Total Steps taken per day", col = "red")

##The mean and median of the total number of steps taken per day
summary(steptable$totalSteps)
```

## What is the average daily activity pattern?

```{r Avg daily activity pattern}

##the average number of steps taken, averaged across all days (y-axis)Second Step per interval

intmean<-with(act,tapply(steps, as.factor(interval), mean,na.rm=T))
intervals<-names(intmean)
avgsteps<-as.numeric(intmean[names(intmean)])
inttable<-data.frame(intervals,avgsteps)

## Plotting the avgsteps across all days w.r.t intervals

plot(intervals,inttable$avgsteps, type="l",col="red")

maxsteps<-max(avgsteps)

##5-minute interval, on average across all the days in the dataset, containing the maximum number of steps

inttable[inttable$avgsteps == maxsteps,]$intervals

```

## Imputing missing values

```{r Imputting missing values}

##total number of missing values in the dataset
sum(as.numeric(is.na(act$steps)))

##inputing missing values with the mean of each day
stepmean<-with(act,tapply(steps, date, mean,na.rm=T))
dates<-names(stepmean)
meanSteps<-as.numeric(stepmean[names(stepmean)])
meanStepTab<-data.frame(dates,meanSteps)

## function to select records with missing value and replacing with the average of that day
createDataSet<-function(act,meanStepTab){
  newact<-act
  for(i in 1:nrow(act)){
    if(is.na(act[i,1])){
      date<-act[i,2]
      tab<-meanStepTab[as.Date(meanStepTab$dates, "%Y-%m-%d") == date,]
      if(is.nan(tab$meanSteps))
        tab$meanSteps=0
      newact[i,1]<-tab$meanSteps
      newact[i,2]<-tab$dates
    }
  }
  newact
}

## New Data set
newact<-createDataSet(act,meanStepTab)

stepsum<-with(newact,tapply(steps, date, sum,na.rm=T))
dates<-names(stepsum)
totalSteps<-as.numeric(stepsum[names(stepsum)])
steptable<-data.frame(dates,totalSteps)

## histogram of the total number of steps taken each day
hist(steptable$totalSteps,main = "Histogram of total steps taken per day",xlab = "Total Steps taken per day", col = "red")

##mean and median total number of steps taken per day
summary(steptable$totalSteps)

```
###As we can see through plot 1 and plot 3 there is no difference from the estimates from the first part of the assignment. There is no impact of imputing missing data on the estimates of the total daily number of steps

## Are there differences in activity patterns between weekdays and weekends?

```{r Activity patterns panel plot}

##Function to Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

dayfun<-function(newact){
  newact$day<-c("")
  for(i in 1:nrow(newact)){
    day<-weekdays(newact[i,2])
    if(day == "Saturday" | day == "Sunday")
      newact[i,4]<-"Weekend"
    else
      newact[i,4]<-"Weekday"
  }
  newact
}

newact<-dayfun(newact)

## Making to sets of data to create the required plot for weekdays and weekends
set1<-subset(newact,day=="Weekend")
set2<-subset(newact,day=="Weekday")

## Finding the average no of steps as per intervals for making the plot

intmean1<-with(set1,tapply(steps, as.factor(interval), mean,na.rm=T))
intmean2<-with(set2,tapply(steps, as.factor(interval), mean,na.rm=T))
intervals<-names(intmean1)
avgsteps<-as.numeric(intmean1[names(intmean1)])
inttable1<-data.frame(intervals,avgsteps,day="Weekend",stringsAsFactors = F)
intervals<-names(intmean2)
avgsteps<-as.numeric(intmean2[names(intmean2)])
inttable2<-data.frame(intervals,avgsteps,day="Weekday",stringsAsFactors = F)

## R Binding both the data sets

resulttable<-rbind(inttable1,inttable2)

## Resultant plot
library(lattice)
xyplot(avgsteps~as.numeric(intervals)|day,data=resulttable,layout=c(1,2),type="l",xlab = "Intervals", ylab ="No of steps average per interval")

```